---
title: "Introduction to Unit Testing in R"
author: "Myfanwy Johnston"
date: "November 7, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What we'll cover today

Basics of unit testing
 - what
 - why
testthat package
assertr package
Exercise

## What we won't cover today

- debugging (2/3 of development)
- 
- Bullet 3

## Except to say

 - Debugging is recommended when the return value (e.g. of a unit test) is erroneous and the error itself is not obvious (University of Freiburg slides)
 - Debugging strategy: 
    * realize you have a bug
    * reproduce/generate the input values that cause the bug
    * isolate the flawed component with a binary search
    * fix it
    * confirm its' successful resolution using the previous input
 - when using unit testing, that last step is really creating an automated test

## Debugging with print()
```{r cars, echo = TRUE}

n = 5
k = 3
power = 0
for (i in 0:k) {
  power <- power * i
  print(power)
}
power

```

# Unit Testing


## Basics

- unit tests focus on the lowest level of a program: very small chunks of code
   - at the function or class level
 - Rmd running after you've changed the code would be an example of larger, informal, integrative test
 
## Main use cases for unit testing

 - ensure that code matches specification
 - detect bugs from changing or adding new code
 
## Characteristics of unit testing

 - each unit test usually consists of multiple simple comarisons
 - focus on boundary issues of parameters
 - quick runtimes that allow automated checks after each code change
 - common quality metric is code coverage
 - checks against pre-defined, expected outcomes
 
## Unit Testing in R

1. Store function `f` subject to testing in "`f.R`" file
2. Source that file via `source`
3. Create file `test.f.R` taht contains the tests
4. Write test, e.g.
```{r, echo = TRUE, eval= FALSE}
test_that("Short description", {
expect_equal(sum(1, 2, 3), 6)
})
```
5. Load package `testthat`
6. Run file via `test_file("test.f.R")`, or all files in a directory via `test_dir(...)`
7. Asses results, i.e. failed tests
 

## Asserts

  Asserts trigger a specific message when a condition is not satisfied.

 - `stop(...)`
 - `stopifnot(...)`
 - package: `asserthat`

  Often used to check type and range of input to functions

## assert example
```{r cube, eval = FALSE, echo=TRUE}

cube_root <- function(x) {
if (class(x) != "numeric") {
stop("Wrong variable class: not a single number")
}
if (x < 0) {
stop("Wrong range: cannot be less than 0")
}
if (!is.finite(x)) {
stop("Wrong range: cannot be infinite or NA")
}
return(x^(1/3))
}
cube_root("error") # should throw an error


```

